Natural Selection Across Three Generations of Americans

### 2024-03-04 response-to-reviewers-revision-1.qmd

When we examine relationships across scores, we indeed test for
significance, always by bootstrapping the sample.

*The authors state that the correlation between the PGI and fertility
quantifies natural selection. However, to my knowledge it is the
covariance following the Robert-Price Identity: Morrissey, M. B., L. E.
B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's
equation in observational studies of natural populations." Journal of
evolutionary biology 23.11 (2010): 2277-2288.*

We quantify natural selection with the regression coefficient of the PGI
on relative lifetime reproductive success. At times in the original
manuscript this was wrongly referred to as a correlation but thus has
now been updated. The polygenic scores are standardized to have a
standard deviation of 1. Under a univariate model, the regression
coefficient is the covariance divided by the variance in the independent
variable. The variance is standardized to 1, so the regression
coefficient is equal to the covariance and thus the two are
interchangeable. A brief discussion of the issue is now given in the
Data section.

In the previous manuscript we used a multivariate model, controlling for
genetic principal components. We now no longer control for the PCs. Now
we residualise the polygenic scores on the PCs before standardizing the
polygenic scores. This ensures our regression estimates are exactly the
covariance between our polygenic scores and RLRS.

@@

> Whilst past research has focused on \[cognitive\] traits, we find
> selection differentials are often of similar magnitudes for health
> related traits. The selection differential for self-rated health is
> greater than those of education and cognitive performance. The most
> significant, positively selected trait was ADHD, which was also found
> in the UK [@hugh2022human]. Future research should study the health
> and medical implications of natural selection, in addition to its
> social implications.

*What is the data base for the mean number of children of people born in
the same year - is pooling appropriate, as the means may vary
considerably according to ethnicity?*

We thought about this, and now agree that it is better to use mean
number of children of whites and of blacks separately for each ethnic
group:

> 'RLRS is calculated separately by ethnicity. This is not ideal,
> because it treats the ethnicities as separate breeding populations.
> But the alternative of calculating pooled RLRS would effectively be
> estimating natural selection in the whole US population by treating
> whites as representative. We therefore focus on the white US
> population, with the caveat that results from this "one data point"
> may not replicate in other ethnicities or countries.'

### 2024-03-21 response-to-reviewers-revision-1.qmd

---
title: "'Natural Selection Across Three Generations of Americans': response to reviewers"
format: pdf
editor: 
  markdown: 
    wrap: 72
bibliography: bibliography-hrs.bib
---

@@

More specifically, we're interested in the pattern of selection effects
across different PGIs, and how it correlates with the PGI's relationship
to income. So we want to include PGIs with small selection effects,
which we predict to have a low absolute correlation to income. Given the
sample size, these PGIs are very likely to have insignificant individual
selection effects, but we shouldn't ignore them in the cross-PGI
analyses.

When we examine relationships across scores, we indeed test for
significance, always by bootstrapping the sample.

*The authors state that the correlation between the PGI and fertility
quantifies natural selection. However, to my knowledge it is the
covariance following the Robert-Price Identity: Morrissey, M. B., L. E.
B. Kruuk, and Alastair J. Wilson. "The danger of applying the breeder's
equation in observational studies of natural populations." Journal of
evolutionary biology 23.11 (2010): 2277-2288.*

We quantify natural selection with the regression coefficient of the PGI
on relative lifetime reproductive success. At times in the original
manuscript this was wrongly referred to as a correlation but this has
now been updated. The polygenic scores are standardized to have a
standard deviation of 1. Under a univariate model, the regression
coefficient is the covariance divided by the variance in the independent
variable. The variance is standardized to 1, so the regression
coefficient is equal to the covariance and thus the two are
interchangeable. A brief discussion of the issue is now given in the
Data section.

### 2024-06-04 hrs-selection.qmd
Small wording tweaks; add some dependencies to renv

titles <- c(white = "White", black = "Black")

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. We include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants. We focus on these because the sample size is large enough. The appendix reports some basic analyses for the `r n_black` genotyped black participants.

@@

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated separately by ethnicity. This is not ideal, because it treats the ethnicities as separate breeding populations. But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative. We therefore focus on the white US population, with the caveat that results from this "one data point" may not replicate in other ethnicities or countries. The mean number of children of people born in each year was calculated using sampling weights.

@@

```{r}
#| label: stat-bootstrap-eth-cor
#| cache: false
#| cache-lazy: false
#| eval: false

@@

```{r}
#| label: stat-pgicorrect
#| cache: false

@@

```{r}
#| label: stat-bootstrap-true-effects
#| cache: false
#| eval: false

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that in the US, SES maps on to race, so that ethnic differences capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents [blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority; @johnson1979minority; @yang2003big; @maralani2013demography]. And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated. But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Another possibility is that the US cohort was exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.

@@

#rand$variables |> 
#    mutate(
#      Ethnicity = fct_relevel(titles[ethnicity], "White", "Black")
#    ) |> 
#    ggplot(aes(rabyear, fill = Ethnicity)) + 
#      geom_bar(position = "identity") +
#      labs(x = "Birth year", y = "Count") +
#      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
#      theme_minimal() +
#      theme(legend.title = element_blank())

@@

year_counts |>
  mutate(
      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
    ) |> 
  ggplot(aes(x = rabyear, y = n, fill = Ethnicity)) + 
      geom_bar(stat="identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())

@@

year_counts |>
   mutate(
      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
    ) |> 
  ggplot(aes(x = rabyear, y = n, fill = Ethnicity )) + 
      geom_bar(stat="identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())

@@

year_counts |>
  mutate(
      Ethnicity = fct_relevel(titles[eth], "White", "Black"),
    )  |> 
  ggplot(aes(x = yearageokid, y = n, fill = Ethnicity )) + 
      geom_bar(stat="identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c(Black = "orange2", White =
      "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())

>>>>

titles <- c(white = "White participants", black = "Black participants")

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. We include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants. We focus on these because the sample size is large enough. The appendix reports some basic analyses for the `r n_black` genotyped black participants. (Throughout, "white" and "black" refer to participants who (a) self-identified as non-Hispanic and as "White/Caucasian" and "Black/African-American" respectively, and who (b) had principal components of SNP data close to the respective population mean. See @hrspgs2020.)

@@

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by the mean number of children of people born in the same year. RLRS is calculated separately by ethnicity. This is not ideal, because it treats the ethnicities as "separate breeding populations" (i.e. subpopulations who rarely interbreed; black and white people intermarried rarely in the HRS generation, but that is due to laws and norms against intermarriage which have now mostly disappeared.) But the alternative of calculating pooled RLRS would effectively be estimating natural selection in the whole US population by treating whites as representative. We therefore focus on the white US population, with the caveat that results from this "one data point" may not replicate in other ethnicities or countries. The mean number of children of people born in each year was calculated using sampling weights.

@@

```{r}
#| label: stat-bootstrap-eth-cor
#| cache: true
#| cache-lazy: false
#| eval: false

@@

```{r}
#| label: stat-pgicorrect
#| cache: true

@@

```{r}
#| label: stat-bootstrap-true-effects
#| cache: true
#| eval: false

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that in the US, SES maps on to race, so that ethnic differences capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on RLRS is more negative among black than white respondents (black respondents: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`, white respondents: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big, @maralani2013demography). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated. But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Another possibility is that the US cohort was exposed to a smaller welfare state than the UK cohort, since many of them had children before the "Great Society" programs of the 1960s.

@@

#rand$variables |> 
#    mutate(
#      Ethnicity = fct_relevel(titles[ethnicity], "White participants",
#                              "Black participants")
#    ) |> 
#    ggplot(aes(rabyear, fill = Ethnicity)) + 
#      geom_bar(position = "identity") +
#      labs(x = "Birth year", y = "Count") +
#      scale_fill_manual(values = c(Black = "orange2", White = "steelblue4")) +
#      theme_minimal() +
#      theme(legend.title = element_blank())

@@

year_counts |>
  mutate(
      Ethnicity = fct_relevel(titles[eth], "White participants",
                              "Black participants"),
    ) |> 
  ggplot(aes(x = rabyear, y = n, fill = Ethnicity)) + 
      geom_bar(stat="identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c("Black participants" = "orange2",
                                   "White participants" = "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())

@@

year_counts |>
   mutate(
      Ethnicity = fct_relevel(titles[eth], "White participants", 
                              "Black participants"),
    ) |> 
  ggplot(aes(x = rabyear, y = n, fill = Ethnicity )) + 
      geom_bar(stat="identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c("Black participants" = "orange2",
                                   "White participants" = "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())

@@

year_counts |>
  mutate(
      Ethnicity = fct_relevel(titles[eth], "White participants",
                              "Black participants"),
    )  |> 
  ggplot(aes(x = yearageokid, y = n, fill = Ethnicity )) + 
      geom_bar(stat="identity") +
      labs(x = "Birth year", y = "Count") +
      scale_fill_manual(values = c("Black participants" = "orange2",
                                   "White participants" = "steelblue4")) +
      theme_minimal() +
      theme(legend.title = element_blank())
